### 2024-04-10 docs/SOSP 2024/schedule.tex
Add HIR MIR optimization section
Changes to scheduling language section

\section{Scheduling Language}
The goal of the scheduling language is to express the following
\begin{itemize}
  \item The order in which the iteration space over the batch of inputs (batch) and trees in the forest (tree) is to be traversed. Note that there is a reduction over the tree dimension.
  \item Intra or inter tree optimizations that are to be performed on a tree or set of trees (tree walk unrolling, pipelining, SIMDize etc).
\end{itemize}

The reasons to use a scheduling language rather than a hard-coded lowering are as follows
\begin{itemize}
  \item Making the scheduling specification external to the compiler allows us to 
  more easily build auto-schedulers and auto-tuners.
  \item It is very hard to come up with a template loop nest that works for all models 
  (for example, tree sizes may vary across the model making it necessary to iterate 
  different number of trees at different times).
  \item A scheduling language will make writing newer locality optimizations faster since 
  no changes to the compiler infrastructure will be needed.
  \item Adding support for additional hardware targets (GPUs, FPGAs), will be much easier
  with a scheduling language.
\end{itemize}

\subsection{Language Definition}
We broadly have three classes of directives in the language. The first is a set 
of loop nest modifiers that are used to specify the structure of the loop nest to
walk the iteration space. The second is a set of clauses that specify intra and 
inter tree optimizations. Finally, we have a class of attributes that control how 
reductions are performed.

\subsubsection{Loop Modifiers}
There are two special index variables -- \op{batch} and \op{tree}. The clauses modify these index variables or 
index variables derived from these (through the application of clauses).
\begin{itemize}
  \item \textbf{tile}: Tile the passed index variable using a fixed tile size.
  \item \textbf{split}: Split the range of the passed index variable into two parts. The range of the first part is specified
  by an argument.
  \item \textbf{unroll}: Unroll an index completely
  \item \textbf{reorder}: Reorder the specified indices. The specified indices must be successive indices in the current loop nest.
  \item \textbf{specialize}: Generate separate code for each iteration of the specified index variable. This is useful 
  while parallelizing across trees and these trees have different depths.
  \item \textbf{gpuDimension}: Maps the specified index variable to represent a dimension in either the GPU kernel grid or thread block.
\end{itemize}

@@

The following are examples of how the loop modifiers can be used.
\begin{itemize}
  \item The loop order used by XGBoost\cite{XGBoost} is (tree, batch) -- walk one tree for all inputs in the batch before 
  moving to the next tree. The corresponding schedule would be
\begin{lstlisting}[style=c++]
  reorder(tree, batch)
\end{lstlisting}

  \item The below schedule computes 2 trees at a time over the whole batch.
\begin{lstlisting}[style=c++]
  tile(tree, t0, t1, 2)
  reorder(t0, batch, t1)
\end{lstlisting}

  \item If we additionally only want to compute over 4 input rows (rather than the whole batch) for
  every 2 tree, and then move onto the next 2 trees for the same set of inputs, then the schedule is as follows. 
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, 4) 
  tile(tree, t0, t1, 2)
  reorder(b0, t0, b1, t1) 
\end{lstlisting}

\end{itemize}

\subsubsection{Optimizations}
The following clauses provide ways to optimize the inference routine being generated.
\begin{itemize}
  \item \textbf{cache}: Cache the working set of one iteration of the specified loop corresponding
  to this index. This can be specified on either batch or tree loops. Specifying it 
  on a batch loop leads to all rows accessed in a single iteration of the loop 
  being cached. Similarly, specifying it on a tree loop leads to all trees accessed in 
  one iteration of that loop being cached.
  \item \textbf{parallel}: Execute the loop corresponding to this index in parallel.
  \item \textbf{interleave}: Interleave the execution of the tree walks within the current index (must be applied on an inner most index).
  \item \textbf{unrollWalk}: Unroll tree walks at the current index. 
  \item \textbf{peelWalk}: Peel the first n steps of the specified tree walk and don't check for leaves for that number of steps.
\end{itemize}

\subsubsection{Reduction Optimization}
\begin{itemize}
  \item \textbf{atomicReduce}: Use atomic memory operations to accumulate values across 
  parallel iterations of the specified loop. 
  \item \textbf{sharedReduce}: Only applies to GPU compilation. Specifies that intermediate
  results are to be stored in shared memory.
  \item \textbf{vectorReduce}: Use vector instructions with the specified vector width 
  to reduce intermediate values across parallel iterations of the specified loop.
\end{itemize}

\TODO{Add examples for RAPIDS, Tahoe (Maybe show some strategies can be encoded?)}

>>>>

\section{Scheduling Language}
\label{sec:schedule}
As we articulated in Section \ref{sec:intro}, there are several 
different configurations and optimizations strategies for decision 
tree inference. The best ones are significantly different across
models, batch sizes and hardware platforms. Therefore, designing
any one hard-coded lowering strategy is not feasible as this would 
make portable performance impossible. To address this problem, we 
design a scheduling language for \Treebeard{}. The scheduling 
language provides an abstract way to specify loop structure and 
other optimizations as an input to the compiler. Making the 
scheduling specification external to the compiler allows to build 
auto-schedulers and auto-tuners (Section \ref{sec:exploring}).
Using a scheduling language will also significantly simplify 
adding support for new hardware as this will likely require
different locality optimizations and loop transformations.

The goal of \Treebeard{}'s scheduling language is to declaratively
express loop structures and the application of other optimizations 
(tree walk unrolling, tree walk interleaving etc.).

\subsection{Language Definition}
The core construct of \Treebeard{}'s scheduling language is an 
\textbf{\emph{index variable}} which abstractly represents a loop. 
The language then provides directives to manipulate these index 
variables. There are two special index variables -- \op{batch} and
\op{tree} that are used to represent the batch and tree loops and all 
other index variables are derived from these. A schedule derives 
new index variables from these root index variables by applying
directives.

\Treebeard{}'s scheduling language has three classes of directives. The first is a set 
of loop modifiers that are used to specify the structure of the loop nest to
walk the iteration space (Table \ref{Tab:LoopModifiers}). The second is a set of 
directives that enable optimizations on a loop (Table \ref{Tab:Optimizations}). 
Finally, we have a class of attributes that enable reduction specific optimizations
(Table \ref{Tab:ReductionOpts}).

% \subsubsection{Loop Modifiers}
% The clauses modify these index variables or 
% index variables derived from these (through the application of clauses).
% \begin{itemize}
%   \item \textbf{tile}: Tile the passed index variable using a fixed tile size.
%   \item \textbf{split}: Split the range of the passed index variable into two parts. The range of the first part is specified
%   by an argument.
%   % \item \textbf{unroll}: Unroll an index completely
%   \item \textbf{reorder}: Reorder the specified indices. The specified indices must be successive indices in the current loop nest.
%   \item \textbf{specialize}: Generate separate code for each iteration of the specified index variable. This is useful 
%   while parallelizing across trees and these trees have different depths.
%   \item \textbf{gpuDimension}: Maps the specified index variable to represent a dimension in either the GPU kernel grid or thread block.
% \end{itemize}

\begin{table}[htb]
  \centering
  \resizebox{\linewidth}{!}{
  \begin{tabularx}{\linewidth}{c | l | l}
   \toprule
   \textbf{Directive} & \textbf{Inputs} & \textbf{Description} \\
   \midrule
   \multirow{4}{*}{\texttt{tile}} & \textbf{indexVar}  & \multirow{4}{*}{\parbox{0.55\linewidth}{Tile the loop corresponding to \textbf{indexVar}
   with the specified tile size. Resulting loops will be represented by \textbf{outer} and \textbf{inner}.}} \\
                                               &  \textbf{outer} & \\
                                               &  \textbf{inner} &  \\
                                               &  \textbf{tileSize} &  \\
   \midrule
   \multirow{6}{*}{\texttt{split}} & \textbf{indexVar}  & \multirow{6}{*}{\parbox{0.55\linewidth}{Fiss the loop represented by \textbf{indexVar}
   at iteration \textbf{splitIter}. Resulting loops will be represented by \textbf{first} and \textbf{second}. Returns a maps from nested 
   index variables to new ones created by splitting.}} \\
                                               &  \textbf{first} & \\
                                               &  \textbf{second} &  \\
                                               &  \textbf{splitIter} &  \\
                                               & & \\
                                               & & \\

   \midrule
   \multirow{4}{*}{\texttt{reorder}} & \textbf{indices[]}  & \multirow{4}{*}{\parbox{0.55\linewidth}{Permute loops corresponding to the specified index variables.
   The loops must be perfectly nested in the current loop structure.}} \\
                                               &   & \\
                                               &   &  \\
                                               &   &  \\
   \midrule
   \multirow{4}{*}{\texttt{specialize}} & \textbf{indexVar}  & \multirow{4}{*}{\parbox{0.55\linewidth}{Generate specialized code for each iteration of the loop. 
   Useful when different iterations of a loop need to execute different code.}} \\
                                               &   & \\
                                               &   &  \\
                                               &   &  \\
   \midrule
   \multirow{3}{*}{\texttt{gpuDimension}} & \textbf{indexVar}  & \multirow{3}{*}{\parbox{0.55\linewidth}{Map the passed index variable to a dimension of either the grid
   or thread block.}} \\
                                               & \textbf{gpuDim}  & \\
                                               &   &  \\

  \bottomrule
  \end{tabularx}
  }
  \vskip 5pt
  \caption{\label{Tab:LoopModifiers} List of all the loop modifiers in \Treebeard{}'s scheduling language. We use \emph{index variable}
  and \emph{loop} interchangeably in descriptions for clarity of exposition.}
\end{table}

% The following are examples of how the loop modifiers can be used.
% \begin{itemize}
%   \item The loop order used by XGBoost\cite{XGBoost} is (tree, batch) -- walk one tree for all inputs in the batch before 
%   moving to the next tree. The corresponding schedule would be
% \begin{lstlisting}[style=c++]
%   reorder(tree, batch)
% \end{lstlisting}

%   \item The below schedule computes 2 trees at a time over the whole batch.
% \begin{lstlisting}[style=c++]
%   tile(tree, t0, t1, 2)
%   reorder(t0, batch, t1)
% \end{lstlisting}

%   \item If we additionally only want to compute over 4 input rows (rather than the whole batch) for
%   every 2 tree, and then move onto the next 2 trees for the same set of inputs, then the schedule is as follows. 
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, 4) 
%   tile(tree, t0, t1, 2)
%   reorder(b0, t0, b1, t1) 
% \end{lstlisting}

% \end{itemize}

% \subsubsection{Optimizations}
% The following clauses provide ways to optimize the inference routine being generated.
% \begin{itemize}
%   \item \textbf{cache}: Cache the working set of one iteration of the specified loop corresponding
%   to this index. This can be specified on either batch or tree loops. Specifying it 
%   on a batch loop leads to all rows accessed in a single iteration of the loop 
%   being cached. Similarly, specifying it on a tree loop leads to all trees accessed in 
%   one iteration of that loop being cached.
%   \item \textbf{parallel}: Execute the loop corresponding to this index in parallel.
%   \item \textbf{interleave}: Interleave the execution of the tree walks within the current index (must be applied on an inner most index).
%   \item \textbf{unrollWalk}: Unroll tree walks at the current index. 
%   \item \textbf{peelWalk}: Peel the first n steps of the specified tree walk and don't check for leaves for that number of steps.
% \end{itemize}

\begin{table}[htb]
  \centering
  \resizebox{\linewidth}{!}{
  \begin{tabularx}{\linewidth}{c | l | l}
   \toprule
   \textbf{Directive} & \textbf{Inputs} & \textbf{Description} \\
   \midrule
   \multirow{4}{*}{\texttt{cache}} & \textbf{indexVar}  & \multirow{4}{*}{\parbox{0.55\linewidth}{Cache the working set of one iteration of the specified loop. 
   Cache rows for a batch loop and trees for a tree loop.}} \\
                                               &  &  \\
                                               &  &  \\
                                               &  &  \\
   \midrule
   \multirow{2}{*}{\texttt{parallel}} & \textbf{indexVar}  & \multirow{2}{*}{\parbox{0.55\linewidth}{Execute the iterations of the specified 
   loop in parallel.}} \\
                                               &  & \\

   \midrule                                               
   \multirow{3}{*}{\texttt{interleave}} & \textbf{indexVar}  & \multirow{3}{*}{\parbox{0.55\linewidth}{Interleave tree walks within the specified 
   loop (must be innermost loop).}} \\
                                               &  & \\
                                               &  & \\

   \midrule                                               
   \multirow{3}{*}{\texttt{unrollWalk}} & \textbf{indexVar}  & \multirow{3}{*}{\parbox{0.55\linewidth}{Unroll tree walks at the specified loop 
   for \textbf{unrollDepth} hops. Loop must be an innermost loop.}} \\
                                               & \textbf{unrollDepth} & \\
                                               &  & \\

  \bottomrule
  \end{tabularx}
  }
  \vskip 5pt
  \caption{\label{Tab:Optimizations} List of optimization directives in \Treebeard{}'s scheduling language. 
  We use \emph{index variable} and \emph{loop} interchangeably in descriptions for clarity of exposition.}
\end{table}

% \subsubsection{Reduction Optimization}
% \begin{itemize}
%   \item \textbf{atomicReduce}: Use atomic memory operations to accumulate values across 
%   parallel iterations of the specified loop. 
%   \item \textbf{sharedReduce}: Only applies to GPU compilation. Specifies that intermediate
%   results are to be stored in shared memory.
%   \item \textbf{vectorReduce}: Use vector instructions with the specified vector width 
%   to reduce intermediate values across parallel iterations of the specified loop.
% \end{itemize}

\begin{table}[htb]
  \centering
  \resizebox{\linewidth}{!}{
  \begin{tabularx}{\linewidth}{c | l | l}
   \toprule
   \textbf{Directive} & \textbf{Inputs} & \textbf{Description} \\
   \midrule
   \multirow{3}{*}{\texttt{atomicReduce}} & \textbf{indexVar}  & \multirow{3}{*}{\parbox{0.55\linewidth}{Use atomic memory operations to accumulate values across 
   parallel iterations of the specified loop.}} \\
                                               &  &  \\
                                               &  &  \\
   \midrule
   \multirow{3}{*}{\texttt{sharedReduce}} & \textbf{indexVar}  & \multirow{3}{*}{\parbox{0.55\linewidth}{Specifies that intermediate
   results are to be stored in shared memory (GPU only).}} \\
                                               &  & \\
                                               &  & \\

   \midrule                                               
   \multirow{4}{*}{\texttt{vectorReduce}} & \textbf{indexVar}  & \multirow{4}{*}{\parbox{0.55\linewidth}{Use vector instructions with the specified vector width 
   to reduce intermediate values across parallel iterations of the specified loop.}} \\
                                               & \textbf{width} & \\
                                               &  & \\
                                               &  & \\

  \bottomrule
  \end{tabularx}
  }
  \vskip 5pt
  \caption{\label{Tab:ReductionOpts} List of reduction optimization directives in \Treebeard{}'s scheduling language. 
  We use \emph{index variable} and \emph{loop} interchangeably in descriptions for clarity of exposition.}
\end{table}

%\TODO{Add examples for RAPIDS, Tahoe (Maybe show some strategies can be encoded?)}

@@
